---
title: "Lab 1 VHE"
author: "Victoria Eastman"
date: "September 19, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Initial EDA

Problem statement: 

```{r}
# Import libraries
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(Hmisc))

setwd("/home/victoriaeastman/berkeley/w271/w271_lab1")
data <- read.csv("challenger.csv")
```

```{r}
glimpse(data)
describe(data)

# I'm curious about the value counts for o-ring failures
table(data$O.ring)
```

Initial findings:

  - 23 data points with no missing values for any variables
  - Dependent variable, O.ring, is categorical and takes three values: 0, 1, and 2 representing the number of o-ring failures on space launches. The mean value is 0.3913 which means the data is skewed towards 0 o-ring failures. Futher investigation shows there were 2 flights with 2 o-ring failures, 5 with 1 failure, and 16 with no failures.
  - The explanatory variables are as follows:
  
    + Temp: temperature at launch (degrees F)
    + Pressure: Combustion pression (psi)
  
  
```{r}
par(mfrow = c(3,2))

# histogram of explanatory variables
for (i in 2:4){
  hist(as.numeric(data[,i]), main=paste0("Histogram of ", colnames(data)[i]), xlab=NA)
  hist(as.numeric(log(data[,i])), main=paste0("Histogram of log(", colnames(data)[i], ")"), xlab=NA)
}

```



```{r}
for (i in 2:3){
  print(ggplot(data, aes(factor(O.ring), data[,i])) +
          geom_boxplot(aes(fill = factor(O.ring))) + 
          geom_jitter() +
          ggtitle(paste0(colnames(data)[i], " by Admit")) + 
          theme(plot.title = element_text(lineheight=1, face="bold")))
}
```